A GPU Implementation of Capon for an Active Sonar System

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	16:55:24	(Thu 3.May)

Consider an $M$ element phased array. Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights and delays be  $w_m[n]$ and $\Delta_m[n]$. The beamformer output $z[n]$ is then given as
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n-\Delta_0]\\x_1[n-\Delta_1]\\\vdots\\x_{M-1}[n-\Delta_{M-1}]}.
\end{align}
The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction. The solution to this optimization problem is given as
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},
\end{gather}
where $\1$ is a row vector of ones, and $\R$ is the spatial covariance matrix. Since $\R$ is unknown, it must be estimated. We will do this by performing some degree of both spatial and temporal averaging. We start with the former, by segmenting the array into $M-L-1$ subarrays of length $L$, where all but one element overlap. The data vector $x_l[n]$ from subarray $l$ is then given as
\begin{gather}
\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T.
\end{gather}
Then the average of all these subarrays are computed,
\begin{gather}
\tilde\R[n] = \frac{1}{M-L+1} \sumb{i=0}{M-L+1} \x_{i+k}[n]\x_{i+k}\H[n]
\end{gather}
Finally, 
\begin{gather}
\tilde\R[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \x[n']\x\H['n]
\end{gather}

To ensure robustness in an active sonar setting, we perform subarray averaging in order to decorrelate noise from the signal, and some degree of temporal averaging to retain \gls{DAS}-like speckle statistics. The $(i,j)$'th element of the sample covariance matrix, $\hat r_{ij}$, is then given as
\begin{gather}
r_{ij}[n] = \frac{1}{(M-L+1)(2N_{\text{avg}}+1)}\sumb{k=0}{M-L} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
\end{gather}
where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $L$ is the subarray size, and $2N_{\text{avg}}+1$ is the number of temporal samples to perform averaging over. Finally, to make sure that $\eR$ is well conditioned, we add a fraction $d$ of the total power of $\eR$ to its diagonal\cite{Synnevag2007},
\begin{align}
\eR = \eR + \I \cdot d \cdot \tr\{\eR[n]\}.
\end{align}

@@

Massively parallel hardware such as \glspl{GPU} is  have a potential for accelerating a great many signal processing tasks, but fully exploiting their potential is not always trivial. Modern \gls{GPU} chips may contain hundreds of processing cores, but the penalty of such a design is that the amount of memory available locally for each core is very limited. There is also a considerable latency involved when moving data between the \gls{CPU} and these cores, so assessing various memory access schemes is crucial to achieve optimal performance.

>>>>

Consider an $M$ element phased array. Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights and delays be  $w_m[n]$ and $\Delta_m[n]$. The beamformer output $z[n]$ is then given as
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n-\Delta_0]\\x_1[n-\Delta_1]\\\vdots\\x_{M-1}[n-\Delta_{M-1}]}.
\end{align}
The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction. The solution to this optimization problem is given as
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},
\end{gather}
where $\1$ is a row vector of ones, and $\R$ is the spatial covariance matrix. Since $\R$ is unknown, it must be estimated. We will do this by performing some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics. We start with the former, by segmenting the array into $M-L-1$ subarrays of length $L$, where all but one element overlap. The data vector $x_l[n]$ from subarray $l$ is given as
\begin{gather}
\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T,
\end{gather}
and the spatially averaged sample covariance matrix $\tilde R$ is given as
\begin{gather}
\tilde\R[n] = \frac{1}{M-L+1} \sumb{i=0}{M-L+1} \x_{i+k}[n]\x_{i+k}\H[n].
\end{gather}
Finally, we take the average of $2N_\text{avg}+1$ temporal samples to form the final sample covariance matrix $\eR$,
\begin{gather}
\eR[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \tilde\R[n'].
\end{gather}
Finally, to make sure that $\eR$ is well conditioned, we add a fraction $d$ of the total power of $\eR$ to its diagonal\cite{Synnevag2007},
\begin{align}
\eR = \eR + \I \cdot d \cdot \tr\{\eR[n]\}.
\end{align}

% To ensure robustness in an active sonar setting, we perform subarray averaging in order to decorrelate noise from the signal, and some degree of temporal averaging to retain \gls{DAS}-like speckle statistics. The $(i,j)$'th element of the sample covariance matrix, $\hat r_{ij}$, is then given as
% \begin{gather}
% r_{ij}[n] = \frac{1}{(M-L+1)(2N_{\text{avg}}+1)}\sumb{k=0}{M-L} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
% \end{gather}
% where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $L$ is the subarray size, and $2N_{\text{avg}}+1$ is the number of temporal samples to perform averaging over.

@@

Massively parallel hardware such as \glspl{GPU} have a potential for accelerating many signal processing tasks, but fully exploiting their potential is not always trivial. Modern \gls{GPU} chips may contain hundreds of processing cores, but the penalty of this design is that the amount of memory available locally for each core is very limited. There is also a considerable latency involved when moving data between the \gls{CPU} and these cores, so assessing various memory access schemes is crucial to achieve optimal performance.
